jax-cosmo: End-to-End Differentiable and GPU Accelerated Cosmology Library

### 2022-02-11 paper.tex
test github de jec
More on NUTS

Now, one is faced with the evaluation of the expectations during gradient descent. The solution is to use what is called \textit{elliptical standardisation} or \textit{re-parametrization trick} or \textit{coordinate transformation} (see references in \citep{2015arXiv150603431K}). Let us illustrate the trick in the case where $q(\xi; \lambda) = \mathcal{N}(\xi; \mu, \Sigma)$ ($\lambda=(\mu, \Sigma)$). One defines $S_\lambda$ such that $S_\lambda(\xi)=\zeta$, eg. $\zeta = L^{-1}(\xi-\mu)$ where $L$ is the Cholesky decomposition of $\Sigma=LL^T$. The Jacobien of this distribution is 1, so the cost function reads
\begin{multline}
-\mathcal{L}(\lambda) = \underbrace{\mathbb{E}_{\eta\sim \mathcal{N}(0,I)}\left[ \log p(T^{-1}(S_\lambda^{-1}(\zeta)),\mathcal{D}) + \log |J_{T^{-1}}(S_\lambda^{-1}(\zeta))| \right]}_{model} \\ + \underbrace{\mathbb{H}[q(\xi;\lambda)]}_{guide}
\label{eq-loss-svi-3}
\end{multline}
where $\mathbb{H}(q)$ is the Shannon entropy of the gaussian distribution which is independent of the model, so its gradient can be computed once for all and reused in user model. Then,
to get $\nabla_\lambda \mathcal{L}$, the $\nabla$ operator can be put inside the expectation which leads to
\begin{multline}
-\nabla_\lambda\mathcal{L}(\lambda) = \mathbb{E}_{\zeta\sim \mathcal{N}(0,I)}\left\{
\left[ \nabla_z \log p(z,\mathcal{D}) \times \nabla_\xi[T^{-1}(\xi)] \right. \right. \\
+ \left. \left. \nabla_\xi \log|J_{T^{-1}}(\xi)| \right] \times \nabla_\lambda S_\lambda^{-1}(\zeta)
\right\}
+ \nabla_\lambda \mathbb{H}[q(\xi;\lambda)]
\label{eq-loss-svi-4}
\end{multline}
For the use case of the MVN (auto)guide, one gets
\begin{align}
-\nabla_\mu \mathcal{L} &= \mathbb{E}_{\zeta\sim \mathcal{N}(0,I)}\left\{
\nabla_z \log p(z,\mathcal{D}) \times \nabla_\xi[T^{-1}(\xi)]
+ \nabla_\xi \log|J_{T^{-1}}(\xi)|
\right\} \nn \\
-\nabla_L \mathcal{L} &= \mathbb{E}_{\zeta\sim \mathcal{N}(0,I)}\left\{
\left[ \nabla_z \log p(z,\mathcal{D}) \times \nabla_\xi[T^{-1}(\xi)] \right.\right. \nn\\
&\left. \left. \qquad\qquad\qquad + \nabla_\xi \log|J_{T^{-1}}(\xi)| \right] \times \xi^T 
\right\}+ (L^{-1})^T
\end{align}
The expectations used in the above expressions are computed with $\zeta$ i.i.d samples from $\mathcal{N}(0,1)$ distribution which reveals the "S" of SVI.

\subsection{Hamiltonian Monte Carlo}

@@

A variant of the traditional HMC sampler has been introduced in reference \citep{nuts}.

\subsubsection{Neural Reparametrisation}
If the SVI method can be used as is to get $z$ i.i.d. samples from the $q(z,\lambda^\ast)$ distribution, the \textit{Neural Transport MCMC} method \citep{2014arXiv1412.5492P,2019arXiv190303704H} is an efficient way to boost HMC efficiency, especially in target distribution unfavourable geometry where for instance the leapfrog integration algorithm has to face squeezed join distributions for a subset of variables. From SVI, one obtains a first approximation of the target distribution and we show that the change-of-variable $z=F_\lambda(\zeta)$ is such that
\begin{equation}
q(z;\lambda) \rightarrow q(\zeta;\lambda) \bydef q(F_\lambda(\zeta)) |J_{F_\lambda}(\zeta)|
\end{equation}
which is evaluated with optimal $\lambda^\ast$ according to maximization of the ELBO (eg. $F_\lambda=T^{-1}\circ S^{-1}_\lambda$). So, one can use a HMC sampler (eg. NUTS) based on $p(\zeta;\mathcal{D})$ distribution, initialized with $\zeta$ samples from $q(\zeta;\lambda^\ast)$, to get a Markov Chain of $N$ samples $(\zeta_i)_{i<N}$. Then, from the transformation  $z_i=F_{\lambda^\ast}(\zeta_i)$ one finally obtain a Marcov Chain with $(z_i)_{i<N}$ samples. The method is accessible throw the \verb|NeuTraReparam| in \numpyro\ library.

\subsection{Discussion}

%
%
\subsubsection{DES-Y1 Example}
%
Using the same DES-Y1 simulation as in previous section, we have used the \numpyro\ library to first use a simple Multivariate Normal distribution as \textit{guide} to approximate the true posterior (\textit{model}) to feed a SVI. The Adam optimizer has been used with a learning rate set to $10^{-3}$.  Then, we have used the Neural Transport method to generate Markov Chains of different sizes by the NUTS sampler. The SVI optimization requires 20,000 steps and has been performed in about 2 hours on a V100 GPU. Concerning the NUTS running on the same GPU, the chains have run 200 warm samples, and then 200 samples, 1000 samples on a single chain, and finally we have run 10 chains (vectorized) with 1000 samples each. The elapse time for each of these 3 runs is 50 minutes, 150 minutes and 5 hours, respectively. Notice that the \texttt{max\_tree\_depth} parameter has been set to 5 instead of the default value of 10 to speed up running. The results of the evolution of the contours compared to the direct NUTS sampling lasting 20 hours on the same GPU are shown on the figures \ref{fig_svi_nuts_1}, \ref{fig_svi_nuts_2}, \ref{fig_svi_nuts_3}. The  agreement is also rather good for the highly correlated $(b_i)_{i:1,\dots,5}$ bias parameters as show on figure \ref{fig_svi_nuts_4}. Results upon ESS are shown on Tab.~\ref{tab-ESS-NUTS_SVI-1}. On can see that the NUTS sampling with a \texttt{max\_tree\_depth} parameter of 5 after the SVI and Neutral Parametrisation gives same order of efficiencies (lower by a factor 2 for some parameters) compared to the NUTS sampling with block mass matrix and \texttt{max\_tree\_depth} parameter of 7.

@@

@@

>>>>

Now, one is faced with the evaluation of the expectations during gradient descent. The solution is to use what is called \textit{elliptical standardisation} or \textit{re-parametrization trick} or \textit{coordinate transformation} (see references in \citep{2015arXiv150603431K}). Let us illustrate the trick in the case where $q(\xi; \lambda) = \mathcal{N}(\xi; \mu, \Sigma)$ ($\lambda=(\mu, \Sigma)$). One defines $S_\lambda$ such that $S_\lambda(\xi)=\zeta$, eg. $\zeta = L^{-1}(\xi-\mu)$ where $L$ is the Cholesky decomposition of $\Sigma=LL^T$. The Jacobien of this distribution is 1, so the cost function reads
\begin{multline}
-\mathcal{L}(\lambda) = \underbrace{\mathbb{E}_{\eta\sim \mathcal{N}(0,I)}\left[ \log p(T^{-1}(S_\lambda^{-1}(\zeta)),\mathcal{D}) + \log |J_{T^{-1}}(S_\lambda^{-1}(\zeta))| \right]}_{model} \\ + \underbrace{\mathbb{H}[q(\xi;\lambda)]}_{guide}
\label{eq-loss-svi-3}
\end{multline}
where $\mathbb{H}(q)$ is the Shannon entropy of the Gaussian distribution which is independent of the model, so its gradient can be computed once for all and reused in user model. Then,
to get $\nabla_\lambda \mathcal{L}$, the $\nabla$ operator can be put inside the expectation which leads to
\begin{multline}
-\nabla_\lambda\mathcal{L}(\lambda) = \mathbb{E}_{\zeta\sim \mathcal{N}(0,I)}\left\{
\left[ \nabla_z \log p(z,\mathcal{D}) \times \nabla_\xi[T^{-1}(\xi)] \right. \right. \\
+ \left. \left. \nabla_\xi \log|J_{T^{-1}}(\xi)| \right] \times \nabla_\lambda S_\lambda^{-1}(\zeta)
\right\}
+ \nabla_\lambda \mathbb{H}[q(\xi;\lambda)]
\label{eq-loss-svi-4}
\end{multline}
For the use case of the MVN (auto)guide, one gets
\begin{align}
-\nabla_\mu \mathcal{L} &= \mathbb{E}_{\zeta\sim \mathcal{N}(0,I)}\left\{
\nabla_z \log p(z,\mathcal{D}) \times \nabla_\xi[T^{-1}(\xi)]
+ \nabla_\xi \log|J_{T^{-1}}(\xi)|
\right\} \nn \\
-\nabla_L \mathcal{L} &= \mathbb{E}_{\zeta\sim \mathcal{N}(0,I)}\left\{
\left[ \nabla_z \log p(z,\mathcal{D}) \times \nabla_\xi[T^{-1}(\xi)] \right.\right. \nn\\
&\left. \left. \qquad\qquad\qquad + \nabla_\xi \log|J_{T^{-1}}(\xi)| \right] \times \xi^T 
\right\}+ (L^{-1})^T
\end{align}
The expectations used in the above expressions are computed with $\zeta$ i.i.d samples from $\mathcal{N}(0,1)$ distribution which reveals the "S" of SVI. Even, if one can sample from the optimized  $q(z,\lambda^\ast)$ distribution, we have used SVI as front-end to  a NUTS sampler which is described in section \ref{sec-Neural-Reparametrisation} as \textit{Neural Reparametrisation}.
%
\subsection{Hamiltonian Monte Carlo}

@@

A variant of the traditional HMC sampler has been introduced in reference \citep{nuts}. It aims to help finding a new point  $x_{i+1}$ from  the current $x_i$. Namely, the root algorithm of HMC deals with the \textit{leapfrog} integration which requires to set two main parameters: $\varepsilon$ the step size and $L$ the number of steps. As stated by the authors, NUTS  uses a recursive algorithm to build a set of likely candidate samples that spans a wide swath of the target distribution, stopping automatically when it starts to double back and retrace its steps. So, in a automatic manner the parameter $L$ is adjusted to produce statistically  a new point $x_{i+1}$ independent from $x_i$, at the expense of the increase the number of model calls. The user has to set a new  parameter  (\texttt{max\_tree\_depth}) which gives in power of $2$ the maximal number of model calls a each generation. Both sampler HMC and NUTS are available in the \numpyro\ library: \texttt{NUTS} is inherited from \texttt{HMC} which in turns is inherited from the generic \texttt{MCMC} class.
%
\subsubsection{Neural Reparametrisation}
\label{sec-Neural-Reparametrisation}
%
If the SVI method can be used as is to get $z$ i.i.d. samples from the $q(z,\lambda^\ast)$ distribution, the \textit{Neural Transport MCMC} method \citep{2014arXiv1412.5492P,2019arXiv190303704H} is an efficient way to boost HMC efficiency, especially in target distribution unfavourable geometry where for instance the leapfrog integration algorithm has to face squeezed join distributions for a subset of variables. From SVI, one obtains a first approximation of the target distribution and we show that the change-of-variable $z=F_\lambda(\zeta)$ is such that
\begin{equation}
q(z;\lambda) \rightarrow q(\zeta;\lambda) \bydef q(F_\lambda(\zeta)) |J_{F_\lambda}(\zeta)|
\end{equation}
which is evaluated with optimal $\lambda^\ast$ according to maximization of the ELBO (eg. $F_\lambda=T^{-1}\circ S^{-1}_\lambda$). So, one can use a HMC sampler (eg. NUTS) based on $p(\zeta;\mathcal{D})$ distribution, initialized with $\zeta$ samples from $q(\zeta;\lambda^\ast)$, to get a Markov Chain of $N$ samples $(\zeta_i)_{i<N}$. Then, from the transformation  $z_i=F_{\lambda^\ast}(\zeta_i)$ one finally obtain a Marcov Chain with $(z_i)_{i<N}$ samples. The method is accessible throw the \verb|NeuTraReparam| in \numpyro\ library.

@@

\subsection{Discussion}
Using the same DES-Y1 simulation as in previous section, we have used the \numpyro\ library to first use a simple Multivariate Normal distribution as \textit{guide} to approximate the true posterior (\textit{model}) to feed a SVI. The Adam optimizer has been used with a learning rate set to $10^{-3}$.  Then, we have used the Neural Transport method to generate Markov Chains of different sizes by the NUTS sampler. The SVI optimization requires 20,000 steps and has been performed in about 2 hours on a V100 GPU. Concerning the NUTS running on the same GPU, the chains have run 200 warm samples, and then 200 samples, 1000 samples on a single chain, and finally we have run 10 chains (vectorized) with 1000 samples each. The elapse time for each of these 3 runs is 50 minutes, 150 minutes and 5 hours, respectively. Notice that the \texttt{max\_tree\_depth} parameter has been set to 5 instead of the default value of 10 to speed up running. The results of the evolution of the contours compared to the direct NUTS sampling lasting 20 hours on the same GPU are shown on the figures \ref{fig_svi_nuts_1}, \ref{fig_svi_nuts_2}, \ref{fig_svi_nuts_3}. The  agreement is also rather good for the highly correlated $(b_i)_{i:1,\dots,5}$ bias parameters as show on figure \ref{fig_svi_nuts_4}. Results upon ESS are shown on Tab.~\ref{tab-ESS-NUTS_SVI-1}. On can see that the NUTS sampling with a \texttt{max\_tree\_depth} parameter of 5 after the SVI and Neutral Parametrisation gives same order of efficiencies (lower by a factor 2 for some parameters) compared to the NUTS sampling with block mass matrix and \texttt{max\_tree\_depth} parameter of 7.

@@

%
%
\subsubsection{DES-Y1 Example}
%
